MaBoSS for HPC environments: Implementations of the continuous time Boolean model simulator for large CPU clusters and GPU accelerators

### 2024-05-02 rebuttal.txt

Q: It is stated that synthetic models with up to 1000 nodes were created. The synthetic models themselves are available, however, the description of the chosen models is also important to be mentioned in the paper: I would suggest a table that lists the most important features of the different models (size, nodes, formula sizes, how many different models were actually taken into account, ..) that were created (this table can also include the characteristics of the 3 real-world models to put in perspective). 
A: We included the proposed table in Results section and extended the first paragraph in the Benchmarking Methodology to describe the table fields.

Q: The statistics are described, but how would a user interpret them/make use of them for themselves? What is this shared associative structure that the model is updating? Maybe a diagram would be helpful for visualization?
A: We added a new section Statistics output and visualization, which includes how the statistics output is represented in the code and how a user can interpret and visualize it.

Q: How does the code actually deal with circular steady states?
A: If a model contains circular steady states, simulated trajectories that would indefinitely transition over circular states will eventually reach the maximum time and their simulation stops. Their statistics are further computed without any change. The original paper (Stoll et al.) discusses the ways how to detect circular steady states from the computed statistics, but we think that describing these methods is out of the scope of this paper. Therefore, we included general information about circular steady states in Statistics output and visualization section and referenced the original paper for the reader who wants to know more detailed information.

Q: There needs to be a description about the original code. Most importantly, it is always referred to, as the CPU version. This suggests, that we are talking about a purely sequential code. In the Methodology, however, it is described to be run on a full 32-core node, which implies already some basic parallelization. This should be made clear from the beginning.
A: Indeed, the implementation of the "CPU version" was early on parallelized to work on multiple cores, using POSIX threads. We added a new subsection to the background section (MaBoSS CPU Implementation) to describe this implementation.

### 2024-05-02 Article.tex
better algorithm wip

A Boolean signaling model consists of $n$ nodes that are either active or inactive, gaining values 1 or 0 respectively. The \emph{state} of the whole model is represented by a vector $S$ of $n$ Boolean values where $S_i$ represents the value of the $i$-th node. We denote the set of all possible states as $\mathcal{S} = \{0, 1\}^n$; thus $|\mathcal{S}| = 2^n$.

Interactions in the model are described as transitions between two states. A single state can have multiple transitions to other states with assigned transition probabilities. In turn, a Boolean network is represented as a directed weighted graph $G = (\mathcal{S}, \rho)$, where $\rho: \mathcal{S} \times \mathcal{S} \rightarrow [0, \infty)$ is a transition function generating \emph{transition rates}. For convenience, it holds that:
\[ \rho(S, S') = 0 \iff \text{there is no transition from}\ S\ \text{to}\ S' \]

\subsection{MaBoSS: Markovian Boolean Stochastic Simulator}
MaBoSS simulates the \emph{asynchronous update strategy}, where at most a single node changes its value in each transition --- consequently, a state $S$ can have at most $n$ possible transitions.

To determine the possible transitions, each node follows the \emph{Boolean logic} $\mathcal{B}_i: \mathcal{S} \rightarrow [0, \infty)$, which determines the expected Poisson-process rate of transitioning to the other value. If $\mathcal{B}_i(S) = 0$, then the transition at node $i$ is not allowed in state $S$.
Given this formalization, the simulation can be also viewed as a continuous-time Markov process.

@@

\begin{algorithm}
\caption{A single iteration of the MaBoSS simulation of a trajectory, given the state $S$ and time $t$.}
\label{alg:iter}
\begin{algorithmic}[1]
\Procedure{TrajectorySimulationStep}{S, t}
\State $\rho_1 \gets \mathcal{B}_1(S), \dots, \rho_n \gets \mathcal{B}_n(S)$ \Comment compute transition rates
\State $r \gets \text{random}([0, \sum_{i=1}^n \rho_i))$
\State $i \gets \min_i \sum_{j=1}^{i-1} \rho_j \leq r < \sum_{j=1}^{i} \rho_j$ \Comment select a node for flipping
\State $S'\gets S\ \text{with the $i$-th bit flipped}$
\State $u \gets \text{random}([0, 1])$
\State $\delta t \gets -\frac{\ln u}{\sum_{i=1}^n \rho_i}$
\State \textbf{return} $(S', t + \delta t)$
\EndProcedure
\end{algorithmic}
\end{algorithm}

To obtain a good overview of the Markovian process, we apply the Gillespie stochastic simulation algorithm\cite{gillespie1976general} (or kinetic Monte-Carlo\cite{young1966monte}
) on the Boolean state space. Multiple trajectories are generated and aggregated in compound trajectory statistics. Commonly obtained statistics include:
\begin{itemize}
    \item \emph{Network state probabilities on a time window} --- Trajectory states are divided by their transition times into time windows based on the time intervals specified by a window size. For each window, the probability of each state is computed as the duration spent in the state divided by the window size. The probabilities of the corresponding windows are then averaged across all subtrajectories.
    \item \emph{Final states} --- The last sampled states from the trajectories are used to compute a final state distribution.
    \item \emph{Fixed states} --- All reached fixed points are used to compute a fixed state distribution.
\end{itemize}

@@

@@

\begin{table}[h]
    \begin{tabular}{l|cccc}
        \textbf{Model}     & \textbf{\# Nodes (non-inter.)} & \textbf{\# Traj.}        & \textbf{Avg. formula size} & \textbf{Avg. traj. length} \\
        \midrule
        \textsc{cellcycle} & $10$ $(4)$                     & $1$M                     & $4$                        & $26$                       \\
        \textsc{sizek}     & $87$ $(4)$                     & $1$M                     & $22$                       & $525$                      \\
        \textsc{Montagud}  & $133$ $(3)$                    & $1$M                     & $4$                        & $197$                      \\
        \hline
        \emph{Synthetic}   & $10 - 1000$ $(5)$              & $1\text{M} - 10\text{M}$ & $10 - 100$                 & $100$                      \\
        \bottomrule
    \end{tabular}
    \centering
    \caption{Main features of the synthetic and real-world models used in the benchmarks. It includes the size of models in terms of nodes and non-internal nodes, the number of simulated trajectories, the average formula size measured as the arithmetic mean of present nodes in each formula, and the average length of all simulated trajectories. Note that not all combinations of features for the synthetic model were used in the benchmarks, see the following figures for more details.}
    \label{tab:model_features}
\end{table}

>>>>

A Boolean signaling model consists of $n$ \emph{nodes} that are either active or inactive, gaining binary values 1 or 0 respectively. The \emph{state} of the whole model is represented by a vector $S$ of $n$ binary values where $S_i$ represents the value of the $i$-th node. We denote the set of all possible states as $\mathcal{S} = \{0, 1\}^n$; thus $|\mathcal{S}| = 2^n$.

Interactions in the model are described as transitions between two states. A single state can have multiple transitions to other states with assigned transition probabilities. In turn, a Boolean network is represented as a directed weighted graph $G = (\mathcal{S}, \rho)$, where $\rho: \mathcal{S} \times \mathcal{S} \rightarrow [0, \infty)$ is a transition function generating \emph{transition rates}. These rates define edge weights of $G$, which are used to compute the probability of a transition from state $S$ to $S'$ in the following way:
\begin{equation}
    P(S \rightarrow S') = \frac{\rho(S, S')}{\sum_{S'' \in \mathcal{S}} \rho(S, S'')}.
    \label{eq:transition}
\end{equation}
For convenience, it holds that
\begin{equation}
\rho(S, S') = 0 \iff \text{there is no transition from}\ S\ \text{to}\ S'.
\end{equation}

\subsection{MaBoSS: Markovian Boolean Stochastic Simulator}
MaBoSS simulates the \emph{asynchronous update strategy}, where at most a single node changes its value in each transition. Therefore, there is a transition from $S$ to $S'$ only if it holds that 
\begin{equation}
    \begin{split}
S_j \ne S'_j & \text{ for a given } j \\
S_i = S'_i & \text{ for } i \ne j.
    \end{split}
\end{equation}
Consequently, $S$ can have at most $n$ possible transitions. In programming terms, $S'$ is obtained by flipping the $j$-th bit of $S$.

To determine the possible transition rates, each node follows the \emph{Boolean logic} $\mathcal{B}_i: \mathcal{S} \rightarrow [0, \infty)$, which determines the expected Poisson-process rate of transitioning to the other value. If $\mathcal{B}_i(S) = 0$, then the transition at node $i$ is not allowed in state $S$. 
Given this formalization, the simulation can be also viewed as a continuous-time Markov process.

The main computational part of a Boolean logic is a \emph{Boolean formula} $B: \mathcal{S} \rightarrow \{0, 1\}$, which consists of logical operators (such as \emph{and}, \emph{or}, \emph{xor}, \emph{not}) with nodes as operands. For example, 
\begin{equation}
    B_i(S) = (S_2 \land S_3) \lor S_4 
\end{equation}
is a Boolean formula for node $i$, where $S_2$, $S_3$ and $S_4$ are binary values of nodes $2$, $3$ and $4$, respectively. Using the values of $B(S)$ and $S_i$, $\mathcal{B}_i(S)$ is defined by a simple rule:
\begin{equation}
    \mathcal{B}_i(S) = \begin{cases}
        0 & \text{if } S_i = B_i(S) \\
        r & \text{otherwise}
    \end{cases}
    \label{eq:transition_rate}
\end{equation}

@@

\begin{algorithm}
\caption{A single iteration of the MaBoSS simulation of a trajectory, given the state $S$ and time $t$.}
\label{alg:iter}
\begin{algorithmic}[1]
\Procedure{TrajectorySimulationStep}{S, t}
\State $\rho_1 \gets \mathcal{B}_1(S), \dots, \rho_n \gets \mathcal{B}_n(S)$ \Comment compute transition rates
\State $r \gets \text{random}([0, \sum_{i=1}^n \rho_i))$
\State $i \gets \min_i \sum_{j=1}^{i-1} \rho_j \leq r < \sum_{j=1}^{i} \rho_j$ \Comment select index of the transition node
\State $S'\gets S\ \text{with the $i$-th bit flipped}$
\State $u \gets \text{random}([0, 1])$
\State $\delta t \gets -\frac{\ln u}{\sum_{i=1}^n \rho_i}$ \Comment compute time to the next transition
\State \textbf{return} $(S', t + \delta t)$
\EndProcedure
\end{algorithmic}
\end{algorithm}

To obtain a good overview of the Markovian process, we apply the Gillespie stochastic simulation algorithm~\cite{gillespie1976general} (or kinetic Monte-Carlo~\cite{young1966monte}) on the Boolean state space. Multiple trajectories are generated and aggregated in compound trajectory statistics. Commonly obtained statistics include:
\begin{itemize}
    \item \emph{Network state probabilities on a time window} --- Trajectory states are divided by their transition times into time windows based on the time intervals specified by a window size. For each window, the probability of each state is computed as the duration spent in the state divided by the window size. The probabilities of the corresponding windows are then averaged across all subtrajectories.
    \item \emph{Final states} --- The last sampled states from the trajectories are used to compute a final state distribution.
    \item \emph{Fixed states} --- All reached fixed points are used to compute a fixed state distribution.
\end{itemize}

@@

\begin{table}
    \begin{tabular}{l|cccc}
        \textbf{Model}     & \textbf{\# Nodes (non-inter.)} & \textbf{\# Traj.}        & \textbf{Avg. formula size} & \textbf{Avg. traj. length} \\
        \midrule
        \textsc{cellcycle} & $10$ $(4)$                     & $1$M                     & $4$                        & $26$                       \\
        \textsc{sizek}     & $87$ $(4)$                     & $1$M                     & $22$                       & $525$                      \\
        \textsc{Montagud}  & $133$ $(3)$                    & $1$M                     & $4$                        & $197$                      \\
        \hline
        \emph{Synthetic}   & $10 - 1000$ $(5)$              & $1\text{M} - 10\text{M}$ & $10 - 100$                 & $100$                      \\
        \bottomrule
    \end{tabular}
    \centering
    \caption{The main features of the synthetic and real-world models used in the benchmarks. It includes the size of models in terms of nodes and non-internal nodes, the number of simulated trajectories, the average formula size measured as the arithmetic mean of present nodes in each formula, and the average length of all simulated trajectories. Note that not all combinations of features for the synthetic model were used in the benchmarks, see the following figures for more details.}
    \label{tab:model_features}
\end{table}

@@
